---
title: "SCENIC_GRN"
output: html_document
editor_options: 
  chunk_output_type: console
---

Load all packages
```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Suppress loading messages when building the HTML
suppressPackageStartupMessages({
  library(SCENIC)
  library(AUCell)
  library(RcisTarget)
  library(SCopeLoomR)
  library(KernSmooth)
  library(BiocParallel)
  library(ggplot2)
  library(data.table)
  library(grid)
  library(ComplexHeatmap)
  library(Seurat)

})

options(width=200)
```

Load the Human_chimp integrated Seurat object
```{r}
Hu_Chimp.Integrated = readRDS("C:/Users/sicon/Desktop/Stanford PhD/1 Courses/2021 Winter/STEMREM 205/STEMREM 205 project/Processed data/human_chimp_subset/Hu_Chimp_Integrated_After_UMAP.rds")

# Change assay to RNA, necessary step
DefaultAssay(Hu_Chimp.Integrated) <- "RNA"
```

Subset the Seurat object 
```{r}
# Subset the seurat object based on cluster of interest. I am using the iPSC cluster (cluster 6 as an example here)

# Check some iPSC marker genes, figure out which cluter reprents iPSC
VlnPlot(Hu_Chimp.Integrated, features = c("POU5F1", "NANOG", "SOX2", "PODXL", "CDH1", "KRT18"), pt.size = 0)

# Check UMAP, cluster 6 is the iPSC cluster
Idents(Hu_Chimp.Integrated) <- "seurat_clusters"
Hu_Chimp_cluster14 = subset(Hu_Chimp.Integrated, idents = "14")

# Split cluster6 by species
Idents(Hu_Chimp_cluster14) <- "Species"
Cluster14_Human = subset(Hu_Chimp_cluster14, idents = "human")
Cluster14_Chimp = subset(Hu_Chimp_cluster14, idents = "chimp")

# save the human cluster 6 Seurat object as an rds file
saveRDS(Cluster6_Human, file = "C:/Users/sicon/Desktop/Stanford PhD/1 Courses/2021 Winter/STEMREM 205/STEMREM 205 project/Processed data/SCENIC_GRN/Cluster6_Human.rds")

# save on server
saveRDS(Cluster14_Human, file = "/data1/stemrem205/sicowang/scenic/Cluster14/Cluster14_Human.rds")
saveRDS(Cluster14_Chimp, file = "/data1/stemrem205/sicowang/scenic/Cluster14/Cluster14_Chimp.rds")


# the code in line 57 is for reading it on the server, I keep it for copying and pasting.
Cluster10_Chimp = readRDS('/data1/stemrem205/sicowang/scenic/Cluster10/Cluster10_Chimp.rds')

# Retrieve data in an expression matrix ('counts' stores raw UMI counts, 'data' stores normlized data, and 'scale.data' stores scaled data)
exprMat_Human = GetAssayData(object = Cluster14_Human, slot = "counts")
exprMat_Chimp = GetAssayData(object = Cluster14_Chimp, slot = "counts")

# Convert dgMatrix to a normal matrix
exprMat_Human = as.matrix(exprMat_Human)
exprMat_Chimp = as.matrix(exprMat_Chimp)

saveRDS(exprMat_Chimp, file = "/data1/stemrem205/sicowang/scenic/Cluster14/exprMat_Chimp.rds")
saveRDS(exprMat_Human, file = "/data1/stemrem205/sicowang/scenic/Cluster10/exprMat_Human.rds")

```

Initialize SCENIC settings
```{r}
org <- "hgnc" # or hgnc, or dmel, or mgi
#dbDir <- "cisTarget_databases" # RcisTarget databases location
dbDir <- "/data1/stemrem205/sicowang/scenic/cisTarget_databases" # RcisTarget databases location

myDatasetTitle <- "SCENIC example on cluster14_Chimp" # choose a name for your analysis
data(defaultDbNames)
dbs <- defaultDbNames[[org]]
#dbs["500bp"] = "hg19-500bp-upstream-7species.mc8nr.feather"
#dbs["10kb"]= "hg19-tss-centered-10kb-7species.mc8nr.feather"

# Save and then Transfer dbs_human.rds to the server
saveRDS(dbs, file = "C:/Users/sicon/Desktop/Stanford PhD/1 Courses/2021 Winter/STEMREM 205/STEMREM 205 project/Processed data/SCENIC_GRN/dbs_human.rds")
# Read this dbs.rds on the server
dbs = readRDS("/data1/stemrem205/sicowang/scenic/dbs_human.rds")

scenicOptions_cluster14_chimp <- initializeScenic(org=org, dbDir=dbDir, dbs=dbs, datasetTitle=myDatasetTitle) 
```

# Co-expression network
## Gene filter/selection
```{r}
# (Adjust minimum values according to your dataset)
genesKept <- geneFiltering(exprMat_Chimp, scenicOptions=scenicOptions_cluster14_chimp,
                           minCountsPerGene=3*.01*ncol(exprMat_Chimp),
                           minSamples=ncol(exprMat_Chimp)*.01)
```

Before proceeding to the network inference, check whether any known relevant genes are filtered-out (if any relevant gene is missing, double-check whether the filters are appropiate): 
```{r checkMissingGenes}
interestingGenes <- c("ZEB2", "SOX2", "PAX6")
# any missing?
interestingGenes[which(!interestingGenes %in% genesKept)]
```

We can now **filter the expression matrix** to contain only these `r length(genesKept)` genes. 
This matrix is now ready for the co-expression analysis.
```{r saveFilteredExprMat}
exprMat_filtered <- exprMat_Chimp[genesKept, ]
dim(exprMat_filtered)
```

## Correlation
Calculate the correlation: 
```{r correlationMat, eval=FALSE}
runCorrelation(exprMat_filtered, scenicOptions_cluster14_chimp)
```

## GENIE3 
```{r}
# Optional: add log (if it is not logged/normalized already)
exprMat_filtered <- log2(exprMat_filtered+1) 
saveRDS(exprMat_filtered, file = "/data1/stemrem205/sicowang/scenic/Cluster14/exprMat_filtered.rds")


# Run GENIE3
runGenie3(exprMat_filtered, scenicOptions_cluster14_chimp)

# After finishing running Genie3, save the scenicOptions as an rds file
saveRDS(scenicOptions_cluster14_chimp, file="int/scenicOptions_cluster14_chimp.rds") 
```

# Build and score the GRN (runSCENIC_...)
```{r}
# Optional: log expression (for TF expression plot, it does not affect any other calculation). Please note this step uses the unfiltered expression matrix.
exprMat_Chimp_log <- log2(exprMat_Chimp+1)
dim(exprMat_Chimp)
```

```{r}
# Some basic settings used by the tutorial, no reason why they chose these numbers
scenicOptions_cluster14_chimp@settings$verbose <- TRUE
scenicOptions_cluster14_chimp@settings$nCores <- 10
scenicOptions_cluster14_chimp@settings$seed <- 123

# Do a quick and make sure two human databases are set in the scenicOptions
scenicOptions_cluster14_chimp@settings$dbs


# Downstream step 1 (it takes 3 minutes to complete on the server)
scenicOptions_cluster14_chimp <- runSCENIC_1_coexNetwork2modules(scenicOptions_cluster14_chimp)

# Downstream step 2 (it takes 6 hours to complete on the server). 
# You need to load a pacakge called BiocParallel. Otherwise you will get an error (could not find function "register") when you try to run the code below. To load BiocParallel, type library(BiocParallel)
scenicOptions_cluster14_chimp <- runSCENIC_2_createRegulons(scenicOptions_cluster14_chimp)

# Downstream step 3 (it takes 14 minutes to complete on the server)
# I got an error when running the code below (Error in .AUCell_calcAUC(geneSets = geneSets, rankings = rankings, nCores = nCores, : Valid 'mctype': 'snow' or 'doMC')
# Reason: Somehow there is a bug in the AUCell pacakge, the error was related to nCores. You couldn't set ncores to numbers other than 1. After adjusting nCores=1, no errors come out.

# Set ncores to 1
scenicOptions_cluster14_chimp@settings$nCores <- 1

# SCENIC step 3 (14 min to complete on the server)
scenicOptions_cluster14_chimp <- runSCENIC_3_scoreCells(scenicOptions_cluster14_chimp, exprMat_Chimp_log)

# Change ncores back to 10 after finishing step 3
scenicOptions_cluster14_chimp@settings$nCores <- 10

# To save status
saveRDS(scenicOptions_cluster14_chimp, file="int/scenicOptions_cluster14_chimp.rds") 
```


******************************************************************************************************************************************************
Data Visualization
## GRN: Regulon targets and motifs
**Genes included in the regulons**: 

```{r seeRegulons}
regulons <- loadInt(scenicOptions_cluster14_chimp, "regulons")
regulons[c("ZEB2", "PAX6", "SOX2", "DLX5")]
```

Note than only regulons with 10 genes or more are scored with AUCell (the numbers in brackets in the regulon names indicate the number of genes in the regulon): 
```{r seeRegulons2}
regulons <- loadInt(scenicOptions_cluster14_chimp, "aucell_regulons")
head(cbind(onlyNonDuplicatedExtended(names(regulons))))
```

Details on the TF-target links: For each TF-target pair, the stats from the intermediate steps are summarized in `loadInt(scenicOptions,"regulonTargetsInfo")` 
```{r motifEnrichmentDT}
regulonTargetsInfo <- loadInt(scenicOptions_cluster14_chimp, "regulonTargetsInfo")
tableSubset <- regulonTargetsInfo[TF=="SOX2" & highConfAnnot==TRUE]
viewMotifs(tableSubset, options=list(pageLength=5)) 
```

Alternative tables, showing more or fewer rows/columns could be generated modifiying this code: 

```{r motifEnrichmentDTsubset, eval=FALSE}
motifEnrichment_selfMotifs_wGenes <- loadInt(scenicOptions_cluster14_chimp, "motifEnrichment_selfMotifs_wGenes")
tableSubset <- motifEnrichment_selfMotifs_wGenes[highlightedTFs=="SOX2"]
viewMotifs(tableSubset) 
```


